---
title: "Chapter 9 Building the Regression Model I"
author: "Bryan Goodrich"
date: "`r Sys.Date()`"
output: 
  html_document:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Chapter 9}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

**********
Chapter 9 -- Building the Regression Model I: Model Selection and Validation
================================================================================
**********

```{r set-global-opts, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, tidy=FALSE)
```

Load the data sets
----------------------------------------------------------------------------

```{r}
library(MPV)  # For PRESS wrapper function
library(leaps)  # bestsubsets

data("CH09TA01", package = "ALSM")
data("CH09TA05", package = "ALSM")
```


Input the Surgical Unit Data
----------------------------------------------------------------------------

```{r}
.data <- CH09TA01
names(.data) <- c(paste("X", 1:8, sep = ""), "Y" , "lnY")
data.names <- c("Bloodclot", "Progindex", "Enzyme", "Liver",
                "Age", "Gender", "Alc.Mod", "Alc.Heavy",
                "Survival", "LnSurvival")
```


TABLE 9.1     (p 350)
----------------------------------------------------------------------------
#### Potential Predictor Variables and Response Variable--Surgical Unit Example

```{r}
{
  print(head(.data))
  cat("... \n")
  print(tail(.data))
}
```



Diagnostic Results Not Shown in Book     (p 351)
----------------------------------------------------------------------------

Included here are the stem-and-leaf plots for each of the predictor variables chosen to be used at this time (i.e., the first four). Also included is the scatterplot matrix and correlation matrix.

Since I think the boxplot is more visually telling than the stem-and-leaf plots, I have also included a 2x2 graphics output of the boxplots for each of the predictor variables.

```{r}
with(.data,
{
    stem(X1, 2)
    stem(X2, 4)
    stem(X3, 4)
    stem(X4)
    cor(.data[1:4])
})

pairs(.data[1:4], labels = data.names[1:4])

with(.data, 
{
    par(mfrow = c(2, 2))
    boxplot(X1, main = "Bloodclot")
    boxplot(X2, main = "Progindex")
    boxplot(X3, main = "Enzyme"   )
    boxplot(X4, main = "Liver"    )
})
```


FIGURE 9.2     (p 351)
----------------------------------------------------------------------------
#### Some Preliminary Residual Plots--Surgical Unit Example

```{r}
par(mfrow = c(2, 2))

fit <- lm(Y ~ X1 + X2 + X3 + X4, .data)
plot(fitted(fit), resid(fit), xlab = "Predicted value", ylab = "Residual")
title("(a) Residual Plot for Y")
abline(0, 0)

qqnorm(resid(fit), xlab = "Expected Value", ylab = "Residual", main = "")
qqline(resid(fit))
title("(b) Normal Plot for Y")

fit <- lm(lnY ~ X1 + X2 + X3 + X4, .data)
plot(fitted(fit), resid(fit), xlab = "Predicted value", ylab = "Residual")
title("(c) Residual Plot for lnY")
abline(0, 0)

qqnorm(resid(fit), xlab = "Expected Value", ylab = "Residual", main = "")
qqline(resid(fit))
title("(d) Normal Prot for lnY")
```


FIGURE 9.3     (p 352)
----------------------------------------------------------------------------
#### Scatter Plot Matrix and Correlation Matrix when Response Variable is lnY--Surgical Unit Example

```{r}
cor(model.frame(fit))
pairs(model.frame(fit), labels = data.names[c(10, 1:4)], pch = 18, col = "grey50")
```


TABLE 9.2     (p 353)
----------------------------------------------------------------------------
#### Selected Criteria Values for Model Selection for All Possible Regression Models--Surgical Unit Example

R contains many ways by which to access the relevant selection criteria values. There is also the regsubsets function in the "leaps" library that calculates most of these values for all the subsets you specify. However, if you desired a way to obtain the values for each combination of models, then you could use an algorithm similar to the follow:

```
    models <- list()
    for(i in seq(p)) models[[i]] <- t(combn(1:p, i))
    models <- lapply(models, function(set) {apply(set, 1, function(row) lm(y ~ ., data[row]))})
    unlist(lapply(models, function(set) lapply(set, sumfun)))
```

The requirements are that `p` is the number of parameters to combine, the data needs to be arranged so the first columns are the predictors, `y` names the response variable (use transform on the full model's frame), and `sumfun` names either your own or some other defined function to operate on lm objects. In this case, there are functions for `AIC` (*stats*), `PRESS` (*MPV*), and others. Since the `models` object contains a list of lists of the possible combinations of `p` predictors, the use of `unlist` is to return the raw values as a vector. Then you can `cbind` them into a new dataframe to summarize results. Note, for BIC results set the `AIC` k value to `log(n)`.

Since `regsubsets` provides most of the criteria in `TABLE 9.2`, I will simply make use of that output. I will also provide some plotting with `regsubsets`. Since this procedure will be repeated a few times, I will make a "wrapper" function to generate the table of interest.

The plots of `regsubsets` produces shaded squares, where shaded means that variable is included in the model. See the help pages for plot.regsubsets for details. This approach can be useful when a corresponding table is large.

```{r best, message=FALSE}
best <- function(model, ...) 
{
  subsets <- regsubsets(formula(model), model.frame(model), ...)
  subsets <- with(summary(subsets),
                  cbind(p = as.numeric(rownames(which)), which, rss, rsq, adjr2, cp, bic))
  
  return(subsets)
}  

round(best(fit, nbest = 4), 4)

subsets <- regsubsets(formula(fit), model.frame(fit), nbest = 4)

plot(subsets)
plot(subsets, scale = "Cp")
plot(subsets, scale = "adjr2")
plot(subsets, scale = "r2")
```



FIGURE 9.4     (p 356)
----------------------------------------------------------------------------
#### Plot of Variables Selection Criteria--Surgical Unit Example

The 1:4 is hard-coded into the 'lines' calls. This could be automated with something like this: 

> `seq(unique(x[, 'p']))`

```{r}
x <- best(fit, nbest = 6)

par(mfrow = c(2, 2), pch = 19)

plot(rsq ~ p, x,   xlab = "(a)", ylab = "Rsq", col = "gray50")
lines(1:4, tapply(x[, "rsq"], x[, "p"], max), lwd = 2)

plot(adjr2 ~ p, x, xlab = "(b)", ylab = "Adj Rsq", col = "gray50")
lines(1:4, tapply(x[, "adjr2"], x[, "p"], max), lwd = 2)

plot(cp ~ p, x,    xlab = "(c)", ylab = "Cp", col = "gray50")
lines(1:4, tapply(x[, "cp"], x[, "p"], min), lwd = 2)

plot(bic ~ p, x,   xlab = "(d)", ylab = "BIC", col = "gray50")
lines(1:4, tapply(x[, "bic"], x[, "p"], min), lwd = 2)
```


FIGURE 9.5     (p 362)
----------------------------------------------------------------------------
#### Plot of Variable Selection Criteria with All Eight Predictors--Surgical Unit Example

```{r}
fit <- lm(lnY ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8, .data)
x   <- best(fit, nbest = 20)

par(mfrow = c(2, 2), pch = 19)
plot(rsq ~ p, x, xlab = "(a)", ylab = "Rsq", col = "gray50")
lines(1:8, tapply(x[, "rsq"], x[, "p"], max), lwd = 2)

plot(adjr2 ~ p, x, xlab = "(b)", ylab = "Adj Rsq", col = "gray50")
lines(1:8, tapply(x[, "adjr2"], x[, "p"], max), lwd = 2)

plot(cp ~ p, x, xlab = "(c)", ylab = "Cp", col = "gray50")
lines(1:8, tapply(x[, "cp"], x[, "p"], min), lwd = 2)

plot(bic ~ p, x, xlab = "(d)", ylab = "BIC", col = "gray50")
lines(1:8, tapply(x[, "bic"], x[, "p"], min), lwd = 2)
```



FIGURE 9.6     (p 363)
----------------------------------------------------------------------------
#### Output for "Best" Two Subsets for Each Subset Size--Surgical Unit Example

```{r}
round(best(fit, nbest = 2), 4)  # Fundamentally same info as in TABLE 9.3
```



FIGURE 9.7     (p 366)
----------------------------------------------------------------------------
#### Forward Stepwise Regression Output--Surgical Unit Example

R comes with a number of stepwise functions. The most basic of these is `step` (*base*). There is also `stepAIC` (*MASS*), which provides for a wider range of object classes. Also, there is `stepwise` (*Rcmdr*) that appears to pose an easier interface to using the BIC instead of AIC calculation. As above, you need to define the `k` parameter appropriately to use the BIC in your stepwise regression methods. Additionally, there's `mle.stepwise` (*wle*). We will focus on only using *base* functions here.

If one wanted to run the process themselves, there exists the intuitive functions `add1` and `drop1` that provide for an easy way to accomplish that. They provide RSS and AIC to make comparisons at each step. For repeated use, the `x` parameter can be passed a model matrix, but no checks are done on its validity. Read their help files for details.

```{r}
BIC <- log(nrow(model.frame(fit)))
fit <- lm(lnY ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8, .data)  # Full Model
step(fit, direction = "forward")      # AIC = -160.77 Drop None
step(fit, dir = "forward",  k = BIC)  # BIC = -142.87 Drop None
step(fit, dir = "backward")           # AIC = -163.83 Drop X4 and X7 
step(fit, dir = "backward", k = BIC)  # BIC = -153.41 Drop X4 through X7 -- This model matches the book's search
```



Input the Surgical Unit Validation Data
----------------------------------------------------------------------------

```{r}
# Import the training and validation data sets, respectively
.data <- do.call("rbind.data.frame", list(CH09TA01, CH09TA05))
names(.data) <- c(paste("X", seq(8), sep = ""), "Y", "lnY")

# Add a factor to which data set the observation belongs
.data <- transform(.data, class = gl(2, 54, labels = c("training", "validation")))
```


TABLE 9.5     (p 374)
----------------------------------------------------------------------------
#### Potential Predictor Variables and Response Variable--Surgical Unit Example

```{r}
{
  print(head(.data))
  cat("... \n")
  print(tail(.data))
}
```


TABLE 9.4     (p 373)
----------------------------------------------------------------------------
#### Regression Results for Candidate Models (9.21), (9.22), and (9.23) Based on Model-Building and Validation Data Sets--Surgical Unit Example

Instead of a table this process will create a list object containing the results provided for each of the model fits for each of the data sets.

```{r reg-result, message=FALSE}
newsummary <- function(model)
{
    list('coefs'    = round(t(summary(model)$coef[, 1:2]), 4),
         'criteria' = cbind('SSE'   = anova(model)["Residuals", "Sum Sq"],
                            'PRESS' = PRESS(model),
                            'MSE'   = anova(model)["Residuals", "Mean Sq"],
                            'Rsq'   = summary(model)$adj.r.squared))
}

newsummary(lm(lnY ~ X1 + X2 + X3 + X8,           .data, class == "training"))
newsummary(lm(lnY ~ X1 + X2 + X3 + X8,           .data, class == "validation"))

newsummary(lm(lnY ~ X1 + X2 + X3 + X6 + X8,      .data, class == "training"))
newsummary(lm(lnY ~ X1 + X2 + X3 + X6 + X8,      .data, class == "validation"))

newsummary(lm(lnY ~ X1 + X2 + X3 + X5 + X6 + X8, .data, class == "training"))
newsummary(lm(lnY ~ X1 + X2 + X3 + X5 + X6 + X8, .data, class == "validation"))
```

