---
title: "Chapter 3 Diagnostics and Remedial Measures"
author: "Bryan Goodrich"
date: "`r Sys.Date()`"
output: 
  html_document:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Chapter 3}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

**********
Chapter 3 -- Diagnostics and Remedial Measures
================================================================================
**********

```{r set-global-opts, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, tidy=FALSE)
```

Load the data sets
----------------------------------------------------------------------------

```{r}
library(MASS)
library(car)
library(lmtest)

data("CH01TA01", package = "ALSM")
data("CH03TA01", package = "ALSM")
data("CH03TA04", package = "ALSM")
data("CH03TA07", package = "ALSM")
data("CH03TA08", package = "ALSM")
data("CH03TA10", package = "ALSM")
```


Input the Toluca Company Data
----------------------------------------------------------------------------

```{r}
.data  <- CH01TA01
names(.data) <- c('x', 'y')
fit <- lm(y ~ x, .data)
```


FIGURE 3.1     (p 101)
----------------------------------------------------------------------------
#### Diagnostic Plots for Predictor Variable--Toluca Company Example

```{r}
par(mfrow = c(2, 2), pch = 19)
stem(.data$x, scale=3)  # Printed to screen, not graphics window

dotchart(.data$x, xlab = "Lot Size", main = "(a) Dot Plot")

plot(.data$x, type = "b", lty = 2, xlab = "Run", ylab = "Lot Size")
title("(b) Sequence Plot")

boxplot(.data$x, horizontal = TRUE, xlab = "Lot Size", main = "(d) Box Plot")
```


FIGURE 3.2     (p 104)
----------------------------------------------------------------------------
#### Diagnostic Residual Plots--Toluca Company Example

```{r}
par(mfrow = c(2, 2), pch = 19)

plot(.data$x, resid(fit), xlab = "Lot Size", ylab = "Residual")
title("(a) Residual Plot against x")

plot(resid(fit), type = "b", lty = 2, xlab = "Run", ylab = "Residual")
title("(b) Sequence Plot")

boxplot(resid(fit), horizontal = TRUE, xlab = "Residual")
title("(c) Box Plot")

qqnorm(resid(fit), xlab = "Expected", ylab = "Residual", main = "")
title("(d) Normal Probability Plot")
qqline(resid(fit))
```


Input The Transit Data
----------------------------------------------------------------------------

```{r}
.data <- CH03TA01
names(.data) <- c('y', 'x')
fit <- lm(y ~ x, .data)
```


FIGURE 3.3     (p 105)
----------------------------------------------------------------------------
#### Scatter Plot and Residual Plot Illustrating Nonlinear Regression--Transit Example

```{r}
plot(y ~ x, .data, ylim = c(0, 8), main = "(a) Scatter Plot")
abline(fit)

plot(.data$x, resid(fit), main = "(b) Residual Plot")
abline(0, 0)
```


TABLE 3.1     (p 105)
----------------------------------------------------------------------------
#### Number of Maps Distributed and Increase in Ridership--Transit Example

```{r}
cbind(
  "Increase in Ridership" = .data$y,
  "Maps Distributed"      = .data$x,
  "Fitted Values"         = round(fitted(fit), 2),
  "Residuals"             = round(resid(fit), 2))
```


Input Toluca Company Data And Reset Model
----------------------------------------------------------------------------

```{r}
.data  <- CH01TA01
names(.data) <- c('x', 'y')
fit <- lm(y ~ x, .data)
```


TABLE 3.2     (p 111)
----------------------------------------------------------------------------
#### Residuals and Expected Values under Normality--Toluca Company Example
 
Could obtain mean square error from model directly using `anova(fit)[2, "Mean Sq"]`. Note that MSE can be calculated below as the `SSE / DF`. 

```{r}
# Expected value under normality comes from equation (3.6)
cbind(
  "Residual"                   = round(resid(fit), 2),
  "Rank"                       = rank(resid(fit)),
  "Exp. Value under Normality" = round(sqrt(deviance(fit) / df.residual(fit)) * 
                                       qnorm((rank(resid(fit)) - 0.375) / (nrow(data) + 0.25)), 2))
```


TABLE 3.3     (p 117)
----------------------------------------------------------------------------
#### Calculations for Brown-Forsythe Test for Constancy of Error Variance--Toluca Company Example

This complex function will print out the conclusion from the decision rule regarding this test. All of the respective elements are contained within the return object. If this object is not caught, it will display to the screen as displayed below.

For completeness, the *car* library has a levene test function called `leveneTest`. It requires the data have a factor by which to group the observations. This is performed using `cut` on the independent variable. Thus, we use the test on the response variable grouped by the cut independent variable as demonstrated below. The outcome of interest is the p-value which is greater than our alpha. We fail to reject the null hypothesis of equal variance, the same conclusion the authors derived. By default `leveneTest` uses median values (Brown-Forsythe test), and it uses an F-test instead of the t-test. Unfortunately, there is nothing of interest returned with the test object beyond what is displayed.

```{r message=FALSE}
levene <- function(model, alpha = 0.05)
{
  f <- function(x) 
  {
    within(x, {
      dsqdev = scale(abs(e - median(e)), T, F)^2
      d      = abs(e - median(e))
    })
  }  # end f
  
  data <- model.frame(model)  # Grab the data used in the model
  data <- transform(data,     # Append the residuals and splitting factor
    e = resid(model),
    group = cut(x, 2, labels = LETTERS[1:2]))

  # Split by group factor and add last TABLE 3.3 columns. Notice the syntax for within()
  data.split <- with(data, split(subset(data, select = -group), group))
  data.split <- lapply(data.split, f)

  
  # Define the relevant variables for hypothesis test and return object
  SSd     <- lapply(data.split, function(x) sum(x$dsqdev))
  dbar    <- lapply(data.split, function(x) mean(x$d))
  n       <- c(Total = nrow(data), lapply(data.split, nrow))   # List of n's
  s       <- sqrt( (SSd$A + SSd$B) / (n$Total - 2) )           # sqrt of (3.9a)
  tstar   <- (dbar$A - dbar$B) / (s * sqrt(1/n$A + 1/n$B))     # (3.9)
  tval    <- qt(1 - alpha/2, n$Total - 2)
  p.value <- 2 * pt(-abs(tval), n$Total - 1)                   # = 0.0495
  
  # Print conclusion of the decision rule
  if (abs(tstar) <= tval) {
    print("error variance is constant")
  } else 
    print("error variance is not constant")

  # return split data and defined variables
  data <- list(
    'data'    = data.split,
    'p.value' = p.value,
    'tstar'   = tstar,
    'tval'    = tval,
    'dbar'    = unlist(dbar),
    'SSd'     = unlist(SSd), 
    'n'       = unlist(n),
    's'       = s)

  return(data)
}  # end levene function

levene(fit)
with(.data, leveneTest(y, cut(x, 2)))  # F value is within acceptance region
```


Breusch-Pagan Test     (p 118)
----------------------------------------------------------------------------

R has a method for calculating this contained in the *lmtest* library. It defaults to a studentized BP test, so an extra parameter will be passed.

Note: The authors have retained an error in this 5th edition since the 4th. They claim a p-value of `0.64 > alpha` which is inconsistent with their results and the outcome of the `bptest` command. The true p-value provided appears to be `1 - 0.64 = 0.36`.

```{r message=FALSE}
bptest(fit, student = FALSE)
qchisq(1 - 0.05, 1) # BP falls within the chisq acceptance region
```


Input the Bank Data
----------------------------------------------------------------------------

```{r}
.data <- CH03TA04
names(.data) <- c('x', 'y')
fit <- lm(y ~ x, .data)
```


TABLE 3.4     (p 120)
----------------------------------------------------------------------------
#### Data and Analysis of Variance Table--Bank Example

```{r}
as.table(cbind(
  "Size of Min. Deposit"   = .data$x,
  "Number of New Accounts" = .data$y
))

anova(fit)
```



FIGURE 3.11     (p 121)
----------------------------------------------------------------------------
#### Scatter Plot and Fitted Regression Line--Bank Example

```{r}
plot(y ~ x, .data)
abline(fit)
```


TABLE 3.5     (p 121)
----------------------------------------------------------------------------
#### Data Arranged by Replicate Number and Minimum Deposit--Bank Example

```{r}
tab           <- do.call("cbind", unstack(.data, y ~ x))
tab[2, 4]     <- NA  # The singular "150" level value is duplicated; remove it
tab           <- rbind(tab, colMeans(tab, na.rm = TRUE))
dimnames(tab) <- list(Replicate = c(1:2, "Mean"), Deposit = dimnames(tab)[[2]])
(tab          <- as.table(tab))
```


TABLE 3.6     (p 126)
----------------------------------------------------------------------------
#### General ANOVA Table for Testing Lack of Fit of Simple Linear Regression Function and ANOVA Table--Bank Example

While R has a basic anova table for lm objects, there is a trick to obtain a further decomposition for the lack of fit and pure error. In particular, with a `y ~ x` regression, we can model `y ~ x + factor(x)` to obtain an anova table with the desired values. Also, the *alr3* package contains a function pureErrorAnova that gives a more complete anova table.

For numeric subscripting, the following will be utilized

```
   Number    Rows                      Columns
   1         x         (Regression)    Df
   2         factor(x) (lack of fit)   Sum Sq
   3         Residuals (pure error)    Mean Sq
```

```{r}
fit.aov <- anova(update(fit, . ~ . + factor(x)))
as.table(cbind(
  'SS' = c('SSR'   =     fit.aov[1,   2], 
           'SSE'   = sum(fit.aov[2:3, 2]),
           'SSLF'  =     fit.aov[2,   2],
           'SSPE'  =     fit.aov[3,   2],
           'Total' = sum(fit.aov[1:3, 2])), 

  'Df' = c(              fit.aov[1,   1],
                     sum(fit.aov[2:3, 1]),
                         fit.aov[2,   1],
                         fit.aov[3,   1],
                     sum(fit.aov[1:3, 1])),  

  'MS' = c(              fit.aov[1,   3],
                     sum(fit.aov[2:3, 2]) / sum(fit.aov[2:3, 1]),
                         fit.aov[2,   3],
                         fit.aov[3,   3],
                         NA)  
))
```


Input the Sales Training data
----------------------------------------------------------------------------

```{r}
.data <- CH03TA07
names(.data) <- c('x', 'y')
fit <- lm(y ~ x, .data)
```


TABLE 3.7     (p 130)
----------------------------------------------------------------------------
#### Use of Square Root Transformation of X to Linearize Regression Relation--Sales Training Example

```{r}
(.data <- transform(.data, sqrtx = sqrt(x)))
```


FIGURE 3.14     (p 131)
----------------------------------------------------------------------------
#### Scater Plots and Residual Plots--Sales Training Example

```{r}
par(mfrow = c(2, 2), pch = 19)
plot(y ~ x, .data, xlab = "Days", ylab = "Performance")
title("(a) Scatter Plot")

plot(y ~ sqrt(x), .data, xlab = expression(sqrt(x)), ylab = "Performance")
title(expression(paste("(b) Scatter Plot against ", sqrt(x))))

plot(resid(fit) ~ sqrt(x), .data, xlab = expression(sqrt(x)), ylab = "Residual")
title(expression(paste("(c) Residual Plot against ", sqrt(x))))

qqnorm(resid(fit), xlab = "Expected", ylab = "Residual", main = "")
title("(d) Normal Probability Plot")
```


Input the Plasma Levels data
----------------------------------------------------------------------------

```{r}
.data <- CH03TA08
names(.data) <- c("x", "y", "z")
fit <- lm(z ~ x, .data)
```



TABLE 3.8     (p 133)
----------------------------------------------------------------------------
#### Use of Logarithmic Transformation of Y to Linearize Regression Relation and Stabilize Error Variance--Plasma Levels Example

```{r}
with(.data, data.frame("Age" = x, "Plasma" = y, "Transform" = z))
```


FIGURE 3.16     (p 134)
----------------------------------------------------------------------------
#### Scatter Plots and Residual Plots--Plasma Levels Example

```{r}
par(mfrow = c(2, 2), pch = 19)
plot(y ~ x, .data, xlab = "Age", ylab = "Plasma Level")
title("(a) Scatter Plot")

plot(z ~ x, .data, xlab = "Age")
title("(b) Scatter Plot with Y' = log(Y)")

plot(.data$x, resid(fit)*100, xlab = "Age", ylab = "Residual x 100")
title("(c) Residual Plot against X")
abline(0, 0)

qqnorm(resid(fit), xlab = "Expected Valued", ylab = "Residual", main = "")
title("(d) Normal Probability Plot")
```


TABLE 3.9 and Figure 3.17    (p 136)
----------------------------------------------------------------------------
#### Box-Cox Results--Plasma Levels Example

R has two Box-Cox functions: `boxcox` (*MASS*) and `boxCox` (*car*). They both take a linear model object as their input. They both produce a graph that maximizes the log-likelihood rather than minimize the sum of squares error. The result is the same.

```{r box-cox, message=FALSE, fig.width=10, fig.height=4}
par(mfrow = c(1, 3))
boxcox(lm(y ~ x, .data))
boxCox(lm(y ~ x, .data))   
boxcox.sse <- function(lambda, model)
{
  x  <- model.frame(model)$x
  y  <- model.frame(model)$y
  K2 <- prod(y)^( 1/length(y) )            # (3.36a)
  K1 <- 1 / (lambda * K2^(lambda - 1))     # (3.36b)
  ifelse(lambda != 0,                      # (3.36)
    assign("W", K1 * (y^lambda - 1)),
    assign("W", K2 * log(y)))

  # Deviance = Residual Sum of Squares
  return(deviance(lm(W ~ x)))  
}

lambda <- seq(-2, 2, by = 0.1)
SSE = sapply(lambda, boxcox.sse, lm(y ~ x, .data))
plot(lambda, SSE, type = "l", xlab = expression(lambda))
abline(v = -0.5, lty = 3)
cbind('lambda' = lambda, 'SSE' = SSE)
```


Input the Toluca Company Data
----------------------------------------------------------------------------

```{r}
.data <- CH01TA01
names(.data) <- c('x', 'y')
fit  <- lm(y ~ x, .data)
```


FIGURE 3.19     (p 140)
----------------------------------------------------------------------------
#### Lowess Curve and Confidence Band for Regression Line--Toluca Company Example

R has methods for lowess regression (`loess`) and loess plotting. We will simply make use of these faculties. As stated in the Chapter 2 walk-through we will plot confidence bands using the built-in R function predict. The R function `scatter.smooth` plots both the scatter plot and lowess curve. `FIGURE 3.18` will be ignored since it can be generated in the same manner.

```{r}
with(.data, scatter.smooth(x, y))
title("Lowess Curve and Linear Regression Confidence Bands")

plot(y ~ x, .data, xlab = "Lot Size", ylab = "Hours")
title("Lowess Curve and Linear Regression Confidence Bands")
with(.data, lines(loess.smooth(x, y), col = "red"))

# Gather confidence bands, ordered by x, and add lines to plot
ci <- cbind(model.frame(fit), predict(fit, int = "c"))[order(.data$x), ]
lines(lwr ~ x, ci, col = "blue", lty = "dashed" )
lines(upr ~ x, ci, col = "blue", lty = "dashed" )
```


Input the Plutonium Measurement Data
----------------------------------------------------------------------------

```{r}
.data <- CH03TA10
names(.data) <- c("y", "x")
```


TABLE 3.10     (p 141)
----------------------------------------------------------------------------
#### Basic Data--Plutonium Measurement Example

```{r}
cbind("Plutonium Activity" = .data$x, "Alpha Count Rate" = .data$y)
```


FIGURE 3.20     (p 142)
----------------------------------------------------------------------------
#### Scatter Plot and Lowess Smoothed Curve--Plutonium Measurement Example
Analytically confirm nonconstant variance by the BP test

```{r warning=FALSE}
plot(y ~ x, .data, pch = 19, xlab = "pCi/g", ylab = "#/sec")
with(.data, lines(loess.smooth(x, y)))  # Warnings may occur

c(Stat  = bptest(lm(y ~ x, .data), student = FALSE)$statistic,
  Chisq = pchisq(.95, 1))  # BP falls outside acceptance region: Reject H0
```


FIGURE 3.21 FIGURE 3.22 and FIGURE 3.23     (p 143-5)
----------------------------------------------------------------------------
#### Regression Output and Diagnostic Plots--Plutonium Measurement Example

```{r}
.data <- .data[-24, ]  # Remove outlier: Record 24
.data <- transform(.data, sqrty = sqrt(y), sqrtx = sqrt(x))

# Linear Models Summay and Anova
summary(lm(y ~ x, .data))
anova(lm(y ~ x, .data))
summary(lm(sqrty ~ x, .data))
anova(lm(sqrty ~ x, .data))
summary(lm(sqrty ~ sqrtx, .data))
anova(lm(sqrty ~ sqrtx, .data)) 

pplot <- function(formula, data) 
{
  fit <- lm(formula, data)
  plot(resid(fit) ~ fitted(fit), xlab = "Fitted", ylab = "Residual")
  title("(b) Residual Plot")
  abline(0, 0)
  
  qqnorm(resid(fit), xlab = "Expected", ylab = "Residual", main = "")
  qqline(resid(fit))
  title("(c) Normal Probability Plot")
}

par(mfrow = c(2, 2), pch = 19)
pplot(y ~ x, .data)          # Untransformed Plots (FIGURE 3.21)
pplot(sqrty ~ x, .data)      # Transformed Response Plots (FIGURE 3.22)
pplot(sqrty ~ sqrtx, .data)  # Transformed Response and Predictor Plots (FIGURE 3.23)


par(mfrow = c(1,1))

fit <- lm(sqrty ~ sqrtx, .data)
scatter.smooth(.data$sqrtx, .data$sqrty, main = "(d) Confidence Band", pch = 19,
               xlab = expression(sqrt(x)), ylab = expression(sqrt(y)))

newdata = data.frame(sqrtx = sort(.data$sqrtx))
pp <- predict(fit, newdata, int = "c")
lines(newdata$sqrtx, pp[, 2], col = "blue")
lines(newdata$sqrtx, pp[, 3], col = "blue")
```
