---
title: "Chapter 4 Simultaneous Inferences and Other Topics"
author: "Bryan Goodrich"
date: "`r Sys.Date()`"
output: 
  html_document:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Chapter 4}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

**********
Chapter 4 -- Simultaneous Inferences and Other Topics in Regression Analysis
================================================================================
**********

```{r set-global-opts, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, tidy=FALSE)
```

Load the data sets
----------------------------------------------------------------------------

```{r}
data("CH01TA01", package = "ALSM")
data("CH04TA02", package = "ALSM")
```


Input the Toluca Company Data
----------------------------------------------------------------------------

```{r}
.data  <- CH01TA01
names(.data) <- c('x', 'y')
fit  <- lm(y ~ x, .data)
```


Bonferroni Joint Confidence Intervals     (p 156)
----------------------------------------------------------------------------

As the authors explain, we're looking for `B = t(1 - 0.10/4; 23)` for the 90% joint confidence interval. But notice that the `confint` funtion uses a 95% confidence interval `t(1 - 0.05/2; 23)` that equals `B`. Therefore, all we require in this instance is to use `confint(fit)` to get the correct result.

The reason this works is because the Bonferroni correction is just a modification to the t-statistic. The `confint` function uses the level parameter to produce the `alpha = 0.05` in the above case. Therefore, what we require is a level to turn `1 - 0.05/2` into `1 - 0.1/4`, or a level such that:

> `1 - a/2 = 1 - (1 - level)/2 = 1 - alpha/4,   alpha = 0.1, a = (1 - level)`

The 4, as we shall see in multiple regression, is really just `2g`, where `g` is the number of confidence coefficients in the family, in this case 2. It turns out that a `level = 1 - alpha/g = 1 - 0.1/2` will give us the desired result:

> `1 - a/2 = 1 - (1 - level)/2 = 1 - (1 - (1 - alpha/g))/2 = 1 - alpha/2g`

In this case, `1 - alpha/g = 0.95`, just is the default level of confint.

```{r}
confint(fit, level = (1 - 0.1/2))
```



Simultaneous Estimation of Mean Responses     (p 158)
----------------------------------------------------------------------------
#### The Working-Hotelling Procedure (90% confidence coefficient)

```{r}
ci.wh <- function(model, newdata, alpha = 0.1)
{
  df    <- nrow(model.frame(model)) - length(coef(model))  # 23
  W     <- sqrt( 2 * qf(1 - alpha, 2, df) )                # 2.2580
  ci    <- predict(model, newdata, se.fit = TRUE)   
  x <- cbind(
    'x'   = newdata,
    's'   = ci$se.fit,
    'fit' = ci$fit,
    'lwr' = ci$fit - W * ci$se.fit,
    'upr' = ci$fit + W * ci$se.fit)
  
  return(x)
}

new <- data.frame(x = c(30, 65, 100))
ci.wh(fit, new)
```


Simultaneous Estimation of Mean Responses     (p 159)
----------------------------------------------------------------------------
#### The Bonferroni Procedure (90% confidence coefficient)

As in the above Bonferroni correction, we require `level = (1 - alpha/g)`. For clarity, part of the return object includes `se.fit` (set to true), which are the standard errors and `residual.scale`, which is equal to the positive squart root of the MSE. This will be used in calculations below.

```{r}
predict(fit, new, int = "c", level = (1 - 0.1/nrow(new)), se.fit = TRUE)
```



Simultaneous Prediction Intervals for New Observations     (p 160)
----------------------------------------------------------------------------
#### The Scheffe and Bonferroni Procedures (95% confidence coefficient)

```{r}
ci.sim <- function(model, newdata, type = c("B", "S"), alpha = 0.05)
{
  g  <- nrow(newdata)
  CI <- predict(model, newdata, se.fit = TRUE)
  M  <- ifelse(match.arg(type) == "B",
          qt(1 - alpha / (2*g), model$df),              # B = (4.9a)
          sqrt( g * qf( 1 - alpha, g, model$df)))       # S = (4.8a)

  spred <- sqrt( CI$residual.scale^2 + (CI$se.fit)^2 )  # (2.38) 
  x <- data.frame(
    "x"     = newdata,
    "spred" = spred,
    "fit"   = CI$fit,
    "lower" = CI$fit - M * spred,
    "upper" = CI$fit + M * spred)
  
  return(x)
}

new <- data.frame(x = c(80, 100))
ci.sim(fit, new, type = "S")
ci.sim(fit, new, type = "B")
```


Input the Warehouse data
----------------------------------------------------------------------------

```{r}
.data <- CH04TA02
names(.data) <- c('x', 'y')
```


TABLE 4.2 and FIGURE 4.1     (pp 162-4)
----------------------------------------------------------------------------
#### Scatter Plot and Fitted Regression through Origin--Warehouse Example

Included in the summary and anova output are the values calculated on pages 163 and 164.

```{r}
fit <- lm(y ~ 0 + x, .data)   # alternate way to exclude intercept: lm(y ~ x - 1)
tab <- transform(.data, 
  x   = round(x, 0),
  y   = round(y, 0),
  xy  = round(x * y, 0),
  xsq = round(x * x, 0),
  fit = round(fitted(fit), 2),
  e   = round(resid(fit), 2))

rbind(tab, Total = colSums(tab))

plot(.data$x, fitted(fit), pch = 19, xlab = "Work United Performed", ylab = "Variable Labor Costs")
abline(fit)

# Some of the results from pp 163-4 are included or derivable from these
summary(fit)
anova(fit)
confint(fit)
```
